\documentclass[11pt]{article}

\usepackage{acl}
\usepackage{times}
\usepackage{latexsym}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{microtype}
\usepackage{inconsolata}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{url}

\title{Real-Time Squat Form Assessment Using MediaPipe and Temporal Conv1D Modeling}

\author{Workout Posture AI Team \\
  Open-Source Project \\
  \texttt{workout-posture-ai@project.local}}

\begin{document}
\maketitle

\begin{abstract}
This paper presents Workout Posture AI, a real-time squat form assessment system that combines MediaPipe pose estimation with a temporal Conv1D classifier and rule-based biomechanical checks. The system targets six common squat errors (forward lean, heel rise, knee cave, knees over toes, shallow depth, and uneven weight) and provides immediate visual feedback, rep counting, and quality scoring. We describe the complete pipeline implemented in the repository: pose capture, landmark smoothing, body-size-invariant feature extraction, temporal windowing, a Conv1D classification model, and a rule engine that encodes biomechanics-inspired thresholds. The training pipeline processes labeled squat videos, applies augmentation, and balances classes for robust learning. The resulting system operates on live camera feeds or prerecorded videos, delivering explainable feedback and real-time coaching without external sensors.
\end{abstract}

\section{Introduction}
Proper squat technique is fundamental for performance and injury prevention, yet common deviations such as forward trunk lean, knee valgus, and insufficient depth can increase joint loading and reduce training efficacy. Biomechanical analyses of the squat emphasize the influence of depth, trunk position, and joint alignment on knee and hip loading \cite{escamilla2010biomechanical, hartmann2013analysis, caterisano2002depth}. At the same time, low-cost camera-based feedback systems can increase adherence and accessibility compared to laboratory motion capture. Workout Posture AI addresses this gap with a real-time, camera-only posture assessment tool that detects form errors, counts repetitions, and scores rep quality for immediate coaching.

The project integrates MediaPipe pose estimation \cite{lugaresi2019mediapipe} with a temporal Conv1D classifier inspired by sequence modeling research \cite{bai2018empirical} and overlays rule-based checks grounded in biomechanics literature. The contributions of this work are: (1) a complete end-to-end pipeline for squat analysis implemented in Python; (2) a body-size-invariant feature representation with 15 interpretable features; (3) a hybrid inference approach combining deterministic rules and a learned temporal model; and (4) a real-time user interface with rep counting, quality scoring, and visual joint feedback.

\section{Related Work}
\textbf{Human pose estimation.} Multi-person pose estimation systems such as OpenPose \cite{cao2017realtime} and Convolutional Pose Machines \cite{wei2016cpm} enabled reliable 2D joint detection in unconstrained images. MediaPipe \cite{lugaresi2019mediapipe} and BlazePose \cite{bazarevsky2020blazepose} extended these ideas to efficient, on-device pipelines for real-time inference. Workout Posture AI builds on MediaPipe Pose (33 landmarks) to provide high-frequency body tracking with sufficient accuracy for exercise feedback.

\textbf{Temporal modeling for movement classification.} Temporal convolutional networks and related Conv1D models are effective for sequence modeling and offer lower latency than recurrent architectures \cite{bai2018empirical}. Our approach leverages a compact Conv1D network over short windows of pose features to classify squat form in real time.

\textbf{Squat biomechanics and technique analysis.} Prior work emphasizes the effect of squat depth and trunk position on joint loading and muscle activation \cite{escamilla2010biomechanical, hartmann2013analysis, caterisano2002depth}. These findings inform the rule-based thresholds used in our system (e.g., forward lean, shallow depth, knee valgus), complementing learned classification for robust coaching.

\section{System Overview}
Workout Posture AI is organized into a modular pipeline (Figure~\ref{fig:pipeline} description provided in text). The key modules in the repository are: \texttt{config.py} for thresholds and paths, \texttt{utils.py} for math and smoothing, \texttt{features.py} for feature extraction, \texttt{rules.py} for deterministic form checks, \texttt{model.py} for the Conv1D architecture, \texttt{train.py} for dataset processing and training, \texttt{analyzer.py} for real-time inference and UI overlays, and \texttt{main.py} for CLI entry points. The system accepts video from a webcam, IP camera, or file, performs pose estimation with MediaPipe, computes features, evaluates rules and the ML model, and renders feedback overlays.

\section{Pose Acquisition and Normalization}
Each frame is processed by MediaPipe Pose, yielding 33 normalized landmarks with visibility scores. Landmarks are converted to a list of dictionaries containing \texttt{x}, \texttt{y}, \texttt{z}, and \texttt{visibility}. Frames with insufficient visibility across required joints (shoulders, hips, knees, ankles) are rejected to avoid unstable feedback. Exponential moving average smoothing is applied to both landmark coordinates and visibility scores with a configurable \(\alpha=0.6\); this value was selected empirically in the project configuration to balance jitter reduction with low-latency responsiveness and can be tuned for different camera conditions.

Body dimensions are estimated from normalized coordinates. Body height is defined as the nose-to-ankle vertical distance, and body width as shoulder-to-shoulder horizontal distance, with safe fallbacks for partial visibility. These measurements normalize distances and ratios to achieve body-size invariance, enabling a single model to generalize across users with different anthropometry.

\section{Feature Engineering}
A 15-dimensional feature vector is extracted per frame and shared between training and inference to enforce feature parity. Features capture joint angles, relative distances, and symmetry measures, all normalized by body height or width. Examples include knee angles, trunk vertical angle, hip depth ratio, knee-forward ratio, and lateral hip shift. This representation balances interpretability with predictive power and is described in Appendix~A.

\section{Rule-Based Form Error Detection}
Six biomechanically motivated rule checks provide immediate and explainable feedback:
\begin{itemize}
  \item \textbf{Knee cave:} detects knees moving inward relative to ankles, normalized by body width.
  \item \textbf{Forward lean:} measures trunk angle from vertical, flagging excessive flexion.
  \item \textbf{Heel rise:} compares heel height to a standing baseline to detect lifting.
  \item \textbf{Knees over toes:} uses depth (\(z\)) offsets to identify excessive forward knee travel.
  \item \textbf{Shallow depth:} evaluates hip depth ratio at the lowest squat point.
  \item \textbf{Uneven weight:} checks hip height asymmetry and knee angle imbalance.
\end{itemize}

Thresholds are derived from normalized measurements configured in \texttt{config.py}. The rules are executed only during the squat phase, and persistent violations are aggregated to score rep quality. Appendix~B lists the exact thresholds used in the implementation.

\section{Temporal Conv1D Model}
The learned classifier operates on sliding windows of 15 frames (15 frames \(\times\) 15 features per frame). The model stacks three Conv1D blocks (64, 128, 64 filters) with batch normalization, max pooling, and global average pooling. A dense layer with dropout (0.4) precedes the softmax output over seven classes (correct + six error classes). This architecture captures short-term temporal dynamics (e.g., knee collapse trends) while maintaining low latency for real-time feedback. The design follows evidence that temporal convolutions are competitive for sequence modeling in compact settings \cite{bai2018empirical}.

\section{Training Pipeline}
Training data are collected in \texttt{SQUAT\_VIDEOS/} and labeled via a filename-to-class mapping. The pipeline scans videos, extracts pose features, and builds overlapping windows (50\% overlap) to increase sample count. Frames are subsampled at a configurable rate (default 2) to reduce redundancy. Horizontal mirroring is used for augmentation, with independent baseline heel calibration per mirrored sequence.

Class imbalance is addressed with inverse-frequency weights. The model is trained using categorical cross-entropy with the Adam optimizer, initial learning rate 0.001, batch size 32, and 20\% validation split. Early stopping (patience 15) and learning-rate reduction on plateau stabilize training. Training artifacts (\texttt{squat\_model.keras} and class labels) are saved to the \texttt{models/} directory for reuse in real-time inference.

\section{Real-Time Analysis and User Feedback}
Real-time processing is implemented by the \texttt{SquatAnalyzer} class. A finite-state machine tracks squat phases (WAITING \(\rightarrow\) STANDING \(\rightarrow\) SQUATTING \(\rightarrow\) STANDING) using the hip-knee depth ratio. Each rep accumulates rule violations; issues persisting for at least 10 frames influence rep quality (good/fair/poor). The UI overlays a skeleton with color-coded joints, highlights current issues, displays rep counts and quality summary, and optionally shows ML predictions with confidence scores.

The system supports live camera streams, IP webcams, and prerecorded video files, enabling deployment in home or gym environments. The ability to switch camera sources and reset counters provides practical usability for training sessions.

\section{Evaluation}
The repository ships with a pretrained model reporting 97.4\% validation accuracy, computed from the training pipeline's 20\% validation split and reported in the project README for the bundled model. The repository does not include a published dataset size or per-class breakdown, so this metric should be interpreted as a single summary value rather than a comprehensive evaluation. While the reported accuracy indicates strong separation across the seven classes, the evaluation remains limited by the scope and diversity of the available video dataset. Future work should include cross-subject validation, multi-environment testing, and comparisons with motion-capture ground truth to more rigorously quantify accuracy and biomechanical fidelity.

\section{Discussion and Limitations}
The hybrid design combines interpretability with statistical learning, which is advantageous for safety-critical coaching where users need actionable feedback. However, limitations remain. The system relies on 2D/monocular depth estimates (normalized \(z\) from MediaPipe), which may be less stable than multi-camera or depth sensors. Occlusions or partial visibility can reduce reliability, requiring explicit visibility checks. The dataset used for training is limited to curated squat videos and may not cover broader variations in camera angle, body type, or equipment. These constraints motivate future expansion of training data, sensor fusion, and adaptive personalization.

\section{Conclusion}
Workout Posture AI provides a practical, open-source pipeline for real-time squat assessment using camera-only inputs. By combining MediaPipe pose tracking, body-size-invariant features, rule-based biomechanics checks, and a temporal Conv1D classifier, the system delivers immediate, explainable coaching with minimal hardware requirements. The modular implementation supports continued research on exercise form assessment and can serve as a foundation for more comprehensive movement analysis systems.

\bibliographystyle{acl_natbib}
\bibliography{references}

\appendix
\section{Appendix A: Feature Definitions}
Table~\ref{tab:features} lists the 15 features computed per frame.

\begin{table*}[t]
  \centering
  \small
  \begin{tabular}{ll}
    \toprule
    \textbf{Feature} & \textbf{Definition (normalized)} \\
    \midrule
    left\_knee\_angle & Hip--knee--ankle angle / 180 \\
    right\_knee\_angle & Hip--knee--ankle angle / 180 \\
    back\_angle & Shoulder--hip--knee angle / 180 \\
    left\_knee\_cave\_ratio & (ankle\_x - knee\_x) / body width \\
    right\_knee\_cave\_ratio & (ankle\_x - knee\_x) / body width \\
    hip\_depth\_ratio & (avg ankle\_y - hip\_y) / body height \\
    left\_heel\_delta & (heel\_y - baseline\_heel\_y) / body height \\
    right\_heel\_delta & (heel\_y - baseline\_heel\_y) / body height \\
    shoulder\_z\_diff & |left\_shoulder\_z - right\_shoulder\_z| \\
    torso\_lean\_ratio & |shoulder\_x - hip\_x| / body width \\
    hip\_symmetry & |left\_hip\_y - right\_hip\_y| / body height \\
    knee\_angle\_diff & left\_knee\_angle - right\_knee\_angle \\
    trunk\_vertical\_angle & trunk angle from vertical / 90 \\
    knee\_forward\_ratio & avg knee\_z - avg ankle\_z \\
    hip\_lateral\_shift & (hip\_x - ankle\_x) / body width \\
    \bottomrule
  \end{tabular}
  \caption{Body-size-invariant features used for training and inference.}
  \label{tab:features}
\end{table*}

\section{Appendix B: Rule Thresholds and Rep Parameters}
Table~\ref{tab:thresholds} summarizes the thresholds used by the rule engine and rep state machine.

\begin{table}[h]
  \centering
  \small
  \begin{tabular}{ll}
    \toprule
    \textbf{Parameter} & \textbf{Value} \\
    \midrule
    Knee cave threshold & 0.12 (fraction of body width) \\
    Forward lean threshold & 50\textdegree{} from vertical \\
    Heel rise threshold & 0.04 (fraction of body height) \\
    Knees over toes threshold & 0.15 (depth offset) \\
    Shallow depth threshold & 0.18 (hip depth ratio) \\
    Uneven hip threshold & 0.06 (fraction of body height) \\
    Uneven knee threshold & 25\textdegree{} angle difference \\
    Standing hip-knee ratio & 0.23 \\
    Squat hip-knee ratio & 0.15 \\
    \bottomrule
  \end{tabular}
  \caption{Rule engine thresholds and rep detection parameters.}
  \label{tab:thresholds}
\end{table}

\section{Appendix C: Training Hyperparameters}
The default training configuration is: 100 epochs, batch size 32, validation split 0.2, early stopping patience 15, learning rate 0.001, and subsample rate 2. These parameters are configurable in \texttt{config.py} and were selected to balance training stability with real-time inference constraints.

\end{document}
